Gut microbes and their genes or something else

### 2023-02-17 manuscript/main.tex
add figures

\addbibresource{zotero_library}

% Highlighting
\usepackage{color}
\usepackage{soul}
\usepackage{xurl}
\usepackage{longtable,booktabs,array}
\usepackage{multirow}
\usepackage{graphicx}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{graph_a}
    \caption{An example graph}
    \label{fig:x cubed graph}
\end{figure}

(A) Study design and overview. Stool samples, cognitive assessments and
neuroimaging were collected from participants in an accelerated
longitudinal design. (B) Cognitive function scores are assessed by
different instruments, but may be normalized to a common IQ-like scale.
(C-D) Principal coordinates analysis using Bray-Curtis dissimilarity on
taxonomic profiles demonstrates high beta diversity, with much of the
first axis of variation explained by increasing age and alpha diversity.
Differences in gene function profiles (E) and neuroimaging (PCA based on
the Euclidean distance of brain region volumes) (F) are likewise
dominated by changes as children age. (G) Permutation analysis of
variance (PERMANOVA) of taxonomic profiles, functional profiles
(annotated as UniRef90s) and neuroimaging, against the metadata of
interest. Variations in taxonomic and functional profiles explain a
modest but significant percent of the variation in cognitive development
in children over 18 months of age. (H) Mantel testing of different
microbial feature matrices, shows overlapping but distinct patterns of
variation. Dotted lines in (B) and (C) show 6 and 18 months, which are
used as cut-offs in some following models.

@@

To assess whether individual microbial species were associated with
cognitive function, we fit multivariable linear regression
\cite{mallickMultivariableAssociationDiscovery2021}
to the relative abundance of each species that had at least 15\%
prevalence in a given age group (Figure 2A, N = 92 for 0--120 months, N
= 46 for 0--6 months, N = 97 for 18--120 months). Only \emph{Blautia
wexlerae} was significantly associated (q value = 0.14, β = 0.0015) with
cognitive function in children under 6 months old after adjusting for
age and maternal education (Figure 2B). \emph{B. wexlerae} was
previously shown to be depleted in children with diabetes
\cite{benitez-paezDepletionBlautiaSpecies2020},
and that oral administration of \emph{B. wexlerae} partially
ameliorated weight gain and inflammation from a high-fat diet in a mouse
model of T2D
\cite{hosomiOralAdministrationBlautia2022,liuBlautiaNewFunctional2021}.
In children over 18 months of age,
several microbial species were significantly enriched (q-value
\textless{} 0.20) in children with higher cognitive function scores,
including \emph{Gordonibacter pamelaeae}, which produces the
neuroprotective metabolite urolithin
\cite{gongUrolithinAlleviatesBloodbrain2022,selmaDescriptionUrolithinProduction2014},
\emph{Asaccharobacter celatus} and
\emph{Adelcreutzia equolifaciens}, which produce phytoestrogen-derived
equol
\cite{maruoAdlercreutziaEquolifaciensGen2008,thawornkunoBiotransformationDaidzeinEquol2009},
and the SCFA-producing probiotic
species such as \emph{Eubacterium eligens} and \emph{Faecalibacterium
prausnitzii}
\cite{ghoshMediterraneanDietIntervention2020}
(Figure
2B).

\includegraphics[width=6.5in,height=4.83333in]{media/image4.png}

\textbf{Figure 2 - Taxa and gene functional groups are associated with cognitive function.}

(A) Volcano plot of multivariable linear model results showing the
relationship between individual taxa and cognitive function score in
children over 18 months of age, controlling for age, maternal education,
and sequencing depth. Taxa that were significant after FDR correction (q
\textless{} 0.2) are colored red. (B) For taxa that were significantly
associated with cognitive function, heatmaps of prevalence, mean
relative abundance, and correlation with cognitive function in different
age groups. (C-F) Enrichment plots for selected neuroactive gene sets
used in feature set enrichment analysis (FSEA). (G) summary of FSEA
results across all samples (left) as well as in the under 6 months
(middle) and over 18 months (right) subsets, colored based on the
significance of the association. Markers indicate the individual
correlation of genes within a gene set, and vertical bars represent the
median correlation of that gene set.

Given that different microbial species might occupy the same metabolic
niche in different individuals, we hypothesized that microbial genes
grouped by functional activity would be associated with cognition. To
test this, we performed feature set enrichment analysis (FSEA) on groups
of genes with neuroactive potential
\cite{valles-colomerNeuroactivePotentialHuman2019}
and concurrent cognitive function score (Table 2,
\protect\hyperlink{_i7fd3z19jdvg}{\uline{Figure 2}}C-G) and found that
several metabolic pathways were significantly enriched or depleted in
children with higher cognitive function scores. This was true both when
considering all age groups together, though the enrichment of most
pathways was more pronounced in children under 6 months or over 18
months. For example, genes for degrading the 3-carbon SCFA propionate
were significantly depleted in children with higher cognitive function
scores across all age groups tested (Table 2; propionate degradation I,
under 6 months, enrichment score (E.S.) = -0.542, corrected p-value (q)
= 0.020; over 18 months, E.S. = -0.674, q = 0.041). Interestingly, genes
for propionate synthesis were also significantly depleted in higher
scoring children over 18 months (E.S. -0.676, q = 0.023), as were genes
for synthesizing the 2 carbon SCFA acetate in children under 6 months
old (acetate synthesis I, E.S. = -0.194, q = 0.153; acetate synthesis
II, E.S. = -0.342, q = 0.020; acetate synthesis III, E.S. = -0.31, q =
0.052). SCFAs are produced by anaerobic fermentation of dietary fiber
and have been linked with immune system regulation as well as directly
with brain function
\cite{dalileRoleShortchainFatty2019}.

@@

\includegraphics[width=6.5in,height=5.19444in]{media/image2.png}

\textbf{Figure 3 - Random forest models predict concurrently measured
cognitive function.}

Comparison of RF predictor importance versus linear models for children
between birth and 6 months old (A), and for those older than 18 months
(B). Colors represent whether the species belong to the group of
top-important features that account for 80\% of the cumulative
importance on the RF model, if that species was significant (q
\textless{} 0.2) in linear models, both, or neither. Ranked feature
importance for taxa in RF models for children between birth and 6 months
old (C), and for those older than 18 months (D). Taxa that are important
for RF models both for children under 6 months and for those over 18
months (E), only for children under 6 months (F), and only for children
over 18 months (G)

@@

\includegraphics[width=6.5in,height=5.19444in]{media/image1.png}

\textbf{Figure 4: Microbial feature importance in predicting brain
volumes in children over 18 months of age.}

(A) Average test-set correlations for prediction of MRI segmentation
data from microbiome and demographics on select segments after repeated
cross validation. (B) Heatmap of average individual relative taxa
importances on each brain segment. Importances are reported as
proportions relative to the sum of importances for each model - since
every model is trained with 132 features, and even distribution of
importance would be 0.75\% for each feature. Segments and taxa ordered
by HCA on a select list of species with high ranks on average importances,
and their respective highest-load segments (C).

@@

Another notable importance cluster contains taxa associated with the
basal forebrain, especially the cingulate and the accumbens. This
cluster contains the previously-discussed \emph{B. ovatus} and \emph{B.
uniformis,} but also \emph{Alistipes finegoldii and Streptococcus
salivaris.} Reduction in the nucleus accumbens has been associated with
depression symptoms
\cite{wackerRoleNucleusAccumbens2009},
and increased levels of the \emph{Alistipes} genus have been
observed in patients with depressive disorder
\cite{jiangAlteredFecalMicrobiota2015}.

While these reports are important, simultaneously probing of the gut
metagenome and brain structure relationship is novel.

>>>>

\addbibresource{zotero_lib}

% Highlighting
\usepackage{color}
\usepackage{soul}
\usepackage{xurl}
\usepackage{longtable,booktabs,array}
\usepackage{multirow}
\usepackage{graphicx}
\graphicspath{assets}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{assets/Figure1.png}
    \caption{
        Study design and overview. (\textbf{A}) Stool samples, cognitive assessments and
        neuroimaging were collected from participants in an accelerated
        longitudinal design. (\textbf{B}) Cognitive function scores are assessed by
        different instruments, but may be normalized to a common IQ-like scale.
        (\textbf{C-D}) Principal coordinates analysis using Bray-Curtis dissimilarity on
        taxonomic profiles demonstrates high beta diversity, with much of the
        first axis of variation explained by increasing age and alpha diversity.
        Differences in gene function profiles (\textbf{E}) and neuroimaging (PCA based on
        the Euclidean distance of brain region volumes) (\textbf{F}) are likewise
        dominated by changes as children age. (\textbf{G}) Permutation analysis of
        variance (PERMANOVA) of taxonomic profiles, functional profiles
        (annotated as UniRef90s) and neuroimaging, against the metadata of
        interest. Variations in taxonomic and functional profiles explain a
        modest but significant percent of the variation in cognitive development
        in children over 18 months of age. (\textbf{H}) Mantel testing of different
        microbial feature matrices, shows overlapping but distinct patterns of
        variation. Dotted lines in (B) and (C) show 6 and 18 months, which are
        used as cut-offs in some following models.
    }
    \label{fig:Figure 1}
\end{figure}

@@

To assess whether individual microbial species were associated with
cognitive function, we fit multivariable linear regression
\cite{mallickMultivariableAssociationDiscovery2021}
to the relative abundance of each species that had at least 15\%
prevalence in a given age group (Figure 2A, N = 92 for 0--120 months, N
= 46 for 0--6 months, N = 97 for 18--120 months). Only \emph{Blautia wexlerae}
was significantly associated (q value = 0.14, $\beta$ = 0.0015) with
cognitive function in children under 6 months old after adjusting for
age and maternal education (Figure 2B). \emph{B. wexlerae} was
previously shown to be depleted in children with diabetes
\cite{benitez-paezDepletionBlautiaSpecies2020},
and that oral administration of \emph{B. wexlerae} partially
ameliorated weight gain and inflammation from a high-fat diet in a mouse
model of T2D
\cite{hosomiOralAdministrationBlautia2022,liuBlautiaNewFunctional2021}.
In children over 18 months of age,
several microbial species were significantly enriched (q-value
\textless{} 0.20) in children with higher cognitive function scores,
including \emph{Gordonibacter pamelaeae}, which produces the
neuroprotective metabolite urolithin
\cite{gongUrolithinAlleviatesBloodbrain2022,selmaDescriptionUrolithinProduction2014},
\emph{Asaccharobacter celatus} and
\emph{Adelcreutzia equolifaciens}, which produce phytoestrogen-derived
equol
\cite{maruoAdlercreutziaEquolifaciensGen2008,thawornkunoBiotransformationDaidzeinEquol2009},
and the SCFA-producing probiotic
species such as \emph{Eubacterium eligens} and \emph{Faecalibacterium
prausnitzii}
\cite{ghoshMediterraneanDietIntervention2020}
(Figure 2B).

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{assets/Figure2.png}
    \caption{
        \textbf{Figure 2 - Taxa and gene functional groups are associated with cognitive function.}
        (A) Volcano plot of multivariable linear model results showing the
        relationship between individual taxa and cognitive function score in
        children over 18 months of age, controlling for age, maternal education,
        and sequencing depth. Taxa that were significant after FDR correction (q
        \textless{} 0.2) are colored red. (B) For taxa that were significantly
        associated with cognitive function, heatmaps of prevalence, mean
        relative abundance, and correlation with cognitive function in different
        age groups. (C-F) Enrichment plots for selected neuroactive gene sets
        used in feature set enrichment analysis (FSEA). (G) summary of FSEA
        results across all samples (left) as well as in the under 6 months
        (middle) and over 18 months (right) subsets, colored based on the
        significance of the association. Markers indicate the individual
        correlation of genes within a gene set, and vertical bars represent the
        median correlation of that gene set.
    }
    \label{fig:Figure 2}
\end{figure}

Given that different microbial species might occupy the same metabolic
niche in different individuals, we hypothesized that microbial genes
grouped by functional activity would be associated with cognition. To
test this, we performed feature set enrichment analysis (FSEA) on groups
of genes with neuroactive potential
\cite{valles-colomerNeuroactivePotentialHuman2019}
and concurrent cognitive function score (Table 2, Figure 2C-G) and found that
several metabolic pathways were significantly enriched or depleted in
children with higher cognitive function scores. This was true both when
considering all age groups together, though the enrichment of most
pathways was more pronounced in children under 6 months or over 18
months. For example, genes for degrading the 3-carbon SCFA propionate
were significantly depleted in children with higher cognitive function
scores across all age groups tested (Table 2; propionate degradation I,
under 6 months, enrichment score (E.S.) = -0.542, corrected p-value (q)
= 0.020; over 18 months, E.S. = -0.674, q = 0.041). Interestingly, genes
for propionate synthesis were also significantly depleted in higher
scoring children over 18 months (E.S. -0.676, q = 0.023), as were genes
for synthesizing the 2 carbon SCFA acetate in children under 6 months
old (acetate synthesis I, E.S. = -0.194, q = 0.153; acetate synthesis
II, E.S. = -0.342, q = 0.020; acetate synthesis III, E.S. = -0.31, q = 0.052).
SCFAs are produced by anaerobic fermentation of dietary fiber
and have been linked with immune system regulation as well as directly
with brain function \cite{dalileRoleShortchainFatty2019}.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidt]{assets/Figure3.png}
    \caption{
        \textbf{Figure 3 - Random forest models predict concurrently measured cognitive function.}
        Comparison of RF predictor importance versus linear models for children
        between birth and 6 months old (A), and for those older than 18 months
        (B). Colors represent whether the species belong to the group of
        top-important features that account for 80\% of the cumulative
        importance on the RF model, if that species was significant (q
        \textless{} 0.2) in linear models, both, or neither. Ranked feature
        importance for taxa in RF models for children between birth and 6 months
        old (C), and for those older than 18 months (D). Taxa that are important
        for RF models both for children under 6 months and for those over 18
        months (E), only for children under 6 months (F), and only for children
        over 18 months (G)
    }
    \label{fig:Figure 3}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{graph_a}
    \caption{
        \textbf{Figure 4: Microbial feature importance in predicting brain volumes in children over 18 months of age.}
        (A) Average test-set correlations for prediction of MRI segmentation
        data from microbiome and demographics on select segments after repeated
        cross validation. (B) Heatmap of average individual relative taxa
        importances on each brain segment. Importances are reported as
        proportions relative to the sum of importances for each model - since
        every model is trained with 132 features, and even distribution of
        importance would be 0.75\% for each feature. Segments and taxa ordered
        by HCA on a select list of species with high ranks on average importances,
        and their respective highest-load segments (C).
    }
    \label{fig:x cubed graph}
\end{figure}

@@

Another notable importance cluster contains taxa associated with the
basal forebrain, especially the cingulate and the accumbens. This
cluster contains the previously-discussed \emph{B. ovatus} and \emph{B.
uniformis,} but also \emph{Alistipes finegoldii and Streptococcus
salivaris.} Reduction in the nucleus accumbens has been associated with
depression symptoms
\cite{wackerRoleNucleusAccumbens2009},
and increased levels of the \emph{Alistipes} genus have been
observed in patients with depressive disorder
\cite{jiangAlteredFecalMicrobiota2015}.
While these reports are important, simultaneously probing of the gut
metagenome and brain structure relationship is novel.
